---
title: "Axial Seamount 2023 analysis"
format: html
editor: visual
---

## Set up environment

```{r}
library(tidyverse)
library(readxl)
```

## Make manifest

```{r}
# files <- read.csv("list_of_files_05012025.txt", header = F)
# head(files)
# length(unique(files$V1))
# fullpath <- "/scratch/group/hu-lab/data/tag-seq-data/18s-16s-axial22_23-seqs_2024-07/"
```

Parse sample ids.

```{r}
# parsed <- files %>% 
#   mutate(SAMPLEID = str_replace(V1, "(\\w*?)_R((\\d+)).fastq.gz","\\1")) %>% 
#   separate(SAMPLEID, into = c("AXIAL","YEAR", "LAB_NUM", "DIVE", "LOCATION", "COLLECTION", "SAMPLETYPE", "ID", "FILTER", "TYPE", "PRIMER"), sep = "_", remove = FALSE) %>% 
#   mutate(READ = case_when(
#     grepl("_R1.fastq.gz", V1) ~ "r1",
#     grepl("_R2.fastq.gz", V1) ~ "r2",
#   ))
# parsed
```

Check stats. 90 samples. 180 when we consider both.

```{r}
# length(parsed$SAMPLEID)
# length(unique(parsed$SAMPLEID))
# 90+90
```

Make manifest files. Read 1 and Read 2 need to be side by side. Columns need to be: sample-id, `forward-absolute-filepath`, and `reverse-absolute-filepath`

```{r}
# manifest_ALL <- parsed %>% 
#   pivot_wider(names_from = READ, values_from = V1) %>% 
#   mutate(`sample-id` = SAMPLEID,
#          `forward-absolute-filepath` = paste(fullpath, r1, sep = ""),
#          `reverse-absolute-filepath` = paste(fullpath, r2, sep = "")) 
# # View(manifest_ALL)
# head(manifest_ALL)
# unique(manifest_ALL$SAMPLETYPE)
```

Separate 18S and 16S manifest file. Include controls for both.

```{r}
# prokaryote_manifest <- manifest_ALL %>% 
#   filter(PRIMER == "16S" | SAMPLETYPE == "control") %>% 
#   select(`sample-id`, `forward-absolute-filepath`, `reverse-absolute-filepath`)
# 
# eukaryote_manifest <- manifest_ALL %>% 
#   filter(PRIMER == "18S" | SAMPLETYPE == "control") %>% 
#   select(`sample-id`, `forward-absolute-filepath`, `reverse-absolute-filepath`)
```

```{r}
# Write output as a manifest file
# write.table(prokaryote_manifest, file = "manifest-axial-prok", quote=FALSE,col.names=TRUE,row.names=FALSE,sep="\t")
# 
# # Write output as a manifest file
# write.table(eukaryote_manifest, file = "manifest-axial-euk", quote=FALSE,col.names=TRUE,row.names=FALSE,sep="\t")
```

# QIIME2

Running QIIME2 here: `/scratch/group/hu-lab/qiime-axial-2023_18S_16S`

# Import & compile ASVs

# Set up

```{r}
library(tidyverse)
```

# Import tsv files from QIIME2

```{r}
asv <- read.delim(file = "input-data/deepsea-18s-merged-asv-table-05022025.tsv", sep = "\t", skip = 1)
# head(asv)
# colnames(asv)
```

```{r}
tax <- read.delim(file = "input-data/taxonomy.tsv", sep = "\t")
head(tax)
```

```{r}
metadata_samples <- read_excel("metadata-axial-2023.xlsx", sheet = "metadata") %>% 
  select(-NUM, -OLD_NAME_ARCHIVE)
biotic <- read_excel("metadata-axial-2023.xlsx", sheet = "biotic")
head(metadata_samples)
head(biotic)
```

Compile and isolate Axial Seamount 2023 samples only

```{r}
asv_long_tax <- asv %>% 
  pivot_longer(cols = -c("X.OTU.ID"), names_to = "SAMPLE_NAME", values_to = "SEQUENCE_COUNT") %>% 
  select(Feature.ID = "X.OTU.ID", SAMPLE_NAME, SEQUENCE_COUNT) %>% 
  filter(grepl("AXIAL_23", SAMPLE_NAME)) %>% 
  # separate(SAMPLE_NAME, into = c("axial", "year", "lab_num", "dive_ctd", "sample_name", "collection_method", "sample_type","sample_identifier",	"filter_processing","primer"), sep = "_", remove = FALSE) %>% 
  filter(SEQUENCE_COUNT > 0) %>% 
  left_join(metadata_samples)

head(asv_long_tax)
# unique(asv_long_tax$collection_method)
colnames(asv_long_tax)
```

## Curate taxonomy

```{r}
head(tax)
```

```{r}
# head(asv_compiled$Taxon)
myzo_18s <- c("Perkinsea", "Chrompodellids", "Apicomplexa")
supergroup_list <- c("Stramenopiles", "Rhizaria", "Alveolata", "Opisthokonta", "Ancyromonadida")
supergroup0_list <- c("Archaeplastida", "CRuMs", "Provora", "Cryptista")
##
tax_refine <- tax %>% 
  # select(Taxon) %>% distinct() %>%
  separate(Taxon, into = c("Domain", "Supergroup_0","Supergroup", "Phylum", "Class", "Order" ,"Family", "Genus_spp"), remove = FALSE, sep = ";") %>% 
  filter(Domain == "Eukaryota") %>% 
  mutate(SUPERGROUP_PHYLUM = case_when(
    ## NAs!
    (is.na(Supergroup) & is.na(Supergroup_0)) ~ "Eukaryote-Unannotated",
    (is.na(Supergroup) & !(is.na(Supergroup_0))) ~ paste(Supergroup_0, "Unannotated", sep = "-"),
    (!(is.na(Supergroup)) & is.na(Supergroup_0)) ~ paste(Supergroup, "Unannotated", sep = "-"),
    (is.na(Phylum) & is.na(Supergroup)) ~ paste(Supergroup_0, "Unannotated", sep = "-"),
    Supergroup_0 == "Eukaryota_X" ~ "Eukaryote-Unannotated",
    # Alveolata
    Class == "Syndiniales" ~ paste(Supergroup, Class, sep = "-"),
    Phylum %in% myzo_18s ~ "Alveolata-Myzozoa",
    Phylum == "Dinoflagellata" ~ "Alveolata-Dinophyta",
    (Supergroup == "Alveolata" & grepl("_X", Phylum)) ~ "Alveolata-Unannotated",
    (Supergroup == "Alveolata" & is.na(Phylum)) ~ "Alveolata-Unannotated",
    # Stramenopiles
    (Supergroup == "Stramenopiles" & grepl("_X", Phylum)) ~ "Stramenopiles-Unannotated",
    (Supergroup == "Stramenopiles" & is.na(Phylum)) ~ "Stramenopiles-Unannotated",
    # Rhizaria
    (Supergroup == "Rhizaria" & grepl("_X", Phylum)) ~ "Rhizaria-Unannotated",
    (Supergroup == "Rhizaria" & is.na(Phylum)) ~ "Rhizaria-Unannotated",
    # Amoebozoa
    Supergroup_0 == "Amoebozoa" & is.na(Supergroup) ~ "Amoebozoa-Unannotated",
    Supergroup_0 == "Amoebozoa" ~ paste(Supergroup_0, Supergroup, sep = "-"),
    Supergroup == "Discosea" ~ "Amoebozoa-Lobosa",
    # 
    Supergroup == "Apusomonada" ~ "Apusozoa-Apusomonadida",
    Supergroup == "Breviatea" ~ "Breviatea-Unannotated",
    Supergroup == "Picozoa" ~ "Picozoa",
    Supergroup == "Ancyromonadida" ~ "Ancyromonadida",
    Supergroup_0 == "Excavata" ~ paste("Excavata", Supergroup, sep = "-"),
    Supergroup_0 == "Hacrobia" ~ paste("Hacrobia", Supergroup, sep = "-"),
    #
    Supergroup %in% supergroup_list ~ paste(Supergroup, Phylum, sep = "-"), 
    Supergroup_0 %in% supergroup0_list ~ paste(Supergroup_0, Supergroup, sep = "-"), 
    Supergroup == "Telonemia" ~ "Telonemia"
    ))
  # select(-Genus_spp) %>% distinct()
# View(asv_compiled_wtax)
unique(tax_refine$SUPERGROUP_PHYLUM)
```

## Plot vent environment at Axial Seamount

```{r}
colnames(asv_long_tax)
# head(asv_long_tax)
axial_df_tax <- asv_long_tax %>% 
  group_by(Feature.ID, dive_ctd, sample_name, collection_method, sample_type, filter, primer, location_name, depth_category) %>% 
    summarise(AVG_SEQ = mean(SEQUENCE_COUNT)) %>% 
  ungroup() %>% 
  left_join(tax) %>% 
  left_join(tax_refine) %>% 
  filter(Taxon != "Unassigned")
head(axial_df_tax)
```

```{r}

# "#1abc9c", "#34495e", "#9b59b6", "#fea02f", "#ff6f61", "#189ab4", "#f5bfb0", "#de6600", "#007a7a", "#ee429b", "#d7b6d7", "red", "yellow", "blue"
# 
# tax_color_metaT <- c("#1abc9c", "#34495e", "#9b59b6", "#fea02f", "#ff6f61", "#189ab4", "#f5bfb0", "#de6600", "#007a7a", "#93c47d","#ee429b", "#d7b6d7", "#e71414", "#ffe8a4", "#2986cc", "#bcbcbc", "#444444")
```

```{r}
unique(axial_df_tax$Supergroup)
```

### Plot vents

```{r, fig.height=7, fig.width=9}
axial_df_tax %>% 
  filter(sample_type == "vent") %>%
  filter(filter == "PES02" | filter == "sterivex") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  filter(!is.na(Supergroup)) %>% 
  filter(depth_category != "above vent") %>% 
  group_by(location_name, depth_category, SUPERGROUP_PHYLUM) %>% 
  summarize(ASV_COUNT = n(),
            SEQUENCE_SUM = sum(AVG_SEQ)) %>% 
  ggplot(aes(y = location_name, x = SEQUENCE_SUM)) +
  geom_bar(stat = "identity", position = "fill", color = "black", aes(fill = SUPERGROUP_PHYLUM)) +
  scale_fill_manual(values = c("#1abc9c", "#34495e","#125223", "#9b59b6", "#fea02f", "#ff6f61", "#189ab4", "#eb7e17","#f5bfb0", "#de6600", "#007a7a", "#93c47d","#230b36","#a1eb17","#666330","#ee429b", "#d7b6d7", "#e71414", "#ffe8a4", "#eb655b","#2986cc", "#bcbcbc","#c5cc49","#1b1c1b", "#444444", "#f0f5f1")) +
  theme_classic() +
  facet_grid(rows = vars(depth_category)) +
  theme(axis.text.x = element_text(color = "black", size = 12),
        axis.text.y = element_text(color = "black", size = 12),
        strip.background = element_blank(),
        strip.text = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(y = "Diffuse vent sites", x = "Relative Sequence Abundance")
# ggsave("figures/vent-nsf-report.svg", width = 9, height = 7, device = "svg", limitsize = FALSE)
```

### Plot vent, background, plume

```{r}
unique(axial_df_tax$depth_category)
```

```{r, fig.height=12, fig.width=9}
axial_df_tax %>% 
  filter(filter == "PES02" | filter == "sterivex") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  filter(sample_type != "control") %>% 
  filter(!is.na(Supergroup)) %>% 
  # filter(depth_category != "above vent") %>% 
  group_by(sample_type, location_name, depth_category, SUPERGROUP_PHYLUM) %>% 
  summarize(ASV_COUNT = n(),
            SEQUENCE_SUM = sum(AVG_SEQ)) %>% 
  ggplot(aes(y = location_name, x = SEQUENCE_SUM)) +
  geom_bar(stat = "identity", position = "fill", color = "black", aes(fill = SUPERGROUP_PHYLUM)) +
  scale_fill_manual(values = c("#1abc9c", "#34495e","#125223", "#9b59b6", "#fea02f", "#ff6f61", "#189ab4", "#eb7e17","#f5bfb0", "#de6600", "#007a7a", "#93c47d","#230b36","#a1eb17","#666330","#ee429b", "#d7b6d7", "#e71414", "#ffe8a4", "#eb655b","#2986cc", "#bcbcbc","#c5cc49","#1b1c1b", "#444444", "#f0f5f1")) +
  theme_classic() +
  facet_grid(rows = vars(sample_type, depth_category), space = "free", scales = "free") +
  theme(axis.text.x = element_text(color = "black", size = 12),
        axis.text.y = element_text(color = "black", size = 12),
        strip.background = element_blank(),
        # strip.text = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(y = "", x = "Relative Sequence Abundance")

```

#### Repeat with filters

```{r}
unique(axial_df_tax$depth_category)
```


```{r, fig.height=12, fig.width=12}
axial_df_tax %>% 
  # filter(filter == "PES02" | filter == "sterivex") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  filter(sample_type != "control") %>% 
  filter(sample_type != "above vent") |> 
  filter(!is.na(Supergroup)) %>% 
  # filter(depth_category != "above vent") %>% 
  group_by(sample_type, location_name, depth_category, SUPERGROUP_PHYLUM, filter) %>% 
  summarize(ASV_COUNT = n(),
            SEQUENCE_SUM = sum(AVG_SEQ)) %>% 
  ggplot(aes(y = location_name, x = SEQUENCE_SUM)) +
  geom_bar(stat = "identity", position = "fill", color = "black", aes(fill = SUPERGROUP_PHYLUM)) +
  scale_fill_manual(values = c("#1abc9c", "#34495e","#125223", "#9b59b6", "#fea02f", "#ff6f61", "#189ab4", "#eb7e17","#f5bfb0", "#de6600", "#007a7a", "#93c47d","#230b36","#a1eb17","#666330","#ee429b", "#d7b6d7", "#e71414", "#ffe8a4", "#eb655b","#2986cc", "#bcbcbc","#c5cc49","#1b1c1b", "#444444", "#f0f5f1")) +
  theme_classic() +
  facet_grid(rows = vars(sample_type, depth_category), cols = vars(filter), space = "free", scales = "free") +
  theme(axis.text.x = element_text(color = "black", size = 12),
        axis.text.y = element_text(color = "black", size = 12),
        strip.background = element_blank(),
        # strip.text = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(y = "", x = "Relative Sequence Abundance")

```

```{r}
head(axial_df_tax)
unique(axial_df_tax$sample_type)
unique(axial_df_tax$depth_category)
```

```{r, fig.height=8, fig.width=9}
axial_df_tax %>% 
  filter(filter == "PES02" | filter == "sterivex") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  filter(sample_type != "control") %>% 
  filter(sample_type != "plume") %>% 
  filter(depth_category != "above vent") %>% 
  filter(!is.na(SUPERGROUP_PHYLUM)) %>% 
  filter(!is.na(Supergroup)) %>% 
  # filter(depth_category != "above vent") %>% 
  group_by(sample_type, location_name, depth_category, SUPERGROUP_PHYLUM) %>% 
  summarize(ASV_COUNT = n(),
            SEQUENCE_SUM = sum(AVG_SEQ)) %>% 
  ggplot(aes(y = location_name, x = SEQUENCE_SUM)) +
  geom_bar(stat = "identity", position = "fill", width = 0.8, color = "black", aes(fill = SUPERGROUP_PHYLUM)) +
  scale_fill_manual(values = c("#1abc9c", "#34495e","#125223", "#9b59b6", "#fea02f", "#ff6f61", "#189ab4", "#eb7e17","#f5bfb0", "#de6600", "#007a7a", "#93c47d","#230b36","#a1eb17","#666330","#ee429b", "#d7b6d7", "#e71414", "#ffe8a4", "#eb655b","#2986cc", "#bcbcbc","#c5cc49","#1b1c1b", "#444444", "#f0f5f1")) +
  theme_classic() +
  facet_grid(rows = vars(sample_type, depth_category), space = "free", scales = "free") +
  theme(axis.text.x = element_text(color = "black", size = 12),
        axis.text.y = element_text(color = "black", size = 12),
        strip.background = element_blank(),
        # strip.text = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(y = "", x = "Relative Sequence Abundance")

```
# Compare filter types at species level

```{r}
head(axial_df_tax)
```


```{r, fig.height=32, fig.width=6}
axial_df_tax %>% 
  filter(sample_type == "vent") %>%
  # filter(filter == "PES02" | filter == "sterivex") %>%
  filter(Supergroup != "Opisthokonta") %>% 
  filter(!is.na(Supergroup)) %>% 
  filter(depth_category != "above vent") %>%
  group_by(sample_type, filter, location_name, depth_category) %>% 
    mutate(SEQ_SUM = sum(AVG_SEQ)) |> 
  ungroup() |> 
  group_by(sample_type, filter, location_name, depth_category, SUPERGROUP_PHYLUM, Taxon) %>% 
  summarize(ASV_COUNT = n(),
            SEQUENCE_SUM = sum(AVG_SEQ),
            REL_ABUN = (SEQUENCE_SUM/SEQ_SUM)
            ) %>% 
  ggplot(aes(x = filter, y = Taxon, fill = REL_ABUN)) +
  geom_tile(color = "grey70") +
  theme_classic() +
  facet_grid(cols = vars(location_name), rows = vars(SUPERGROUP_PHYLUM), scale = "free", space = "free") +
  theme(axis.text.x = element_text(color = "black", size = 12),
        axis.text.y = element_blank(), 
        # axis.text.y = element_text(color = "black", size = 12),
        strip.background = element_blank(),
        # strip.text = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank()) +
  labs(y = "Diffuse vent sites", x = "Relative Sequence Abundance")
```

```{r}

```


# Plot metadata

```{r}
head(biotic)
biotic_refined <- biotic %>% 
  select(dive_ctd, sample_name, sample_type, location_name, depth_category, depth, prok) %>% 
  drop_na() %>% 
  group_by(location_name, sample_type, depth_category) %>% 
  summarise(mean_prok = mean(prok),
            max_prok = max(prok),
            min_prok = min(prok))
# View(biotic_refined)
```

```{r, fig.height=5, fig.width=5}
ggplot(biotic_refined, aes(x = mean_prok, y = location_name)) +
  geom_point(stat = "identity", color = "black")+
  geom_errorbar(aes(xmin = min_prok, xmax = max_prok), width = 0.2) +
  scale_x_log10() +
  facet_grid(rows = vars(sample_type, depth_category), space = "free", scales = "free") +
  theme_classic() +
    theme(axis.text.x = element_text(color = "black", size = 12),
        axis.text.y = element_text(color = "black", size = 12),
        strip.background = element_blank(),
        # strip.text = element_blank(),
        legend.position = "bottom",
        legend.title = element_blank(),
        panel.grid.major = element_line(color = "grey")) +
  labs(y = "", x = "Log prokaryotic cells per ml")
```

# Session Info

```{r}
sessionInfo()
```
