---
title: "Axial Seamount 2023 analysis"
format: html
editor: visual
---

## Set up environment

```{r}
library(tidyverse)
```

## Make manifest

```{r}
files <- read.csv("list_of_files_05012025.txt", header = F)
head(files)
length(unique(files$V1))
fullpath <- "/scratch/group/hu-lab/data/tag-seq-data/18s-16s-axial22_23-seqs_2024-07/"
```

Parse sample ids.
```{r}
parsed <- files %>% 
  mutate(SAMPLEID = str_replace(V1, "(\\w*?)_R((\\d+)).fastq.gz","\\1")) %>% 
  separate(SAMPLEID, into = c("AXIAL","YEAR", "LAB_NUM", "DIVE", "LOCATION", "COLLECTION", "SAMPLETYPE", "ID", "FILTER", "TYPE", "PRIMER"), sep = "_", remove = FALSE) %>% 
  mutate(READ = case_when(
    grepl("_R1.fastq.gz", V1) ~ "r1",
    grepl("_R2.fastq.gz", V1) ~ "r2",
  ))
parsed
```

Check stats.
90 samples. 180 when we consider both.
```{r}
length(parsed$SAMPLEID)
length(unique(parsed$SAMPLEID))
# 90+90
```

Make manifest files. Read 1 and Read 2 need to be side by side.
Columns need to be:
sample-id, `forward-absolute-filepath`, and `reverse-absolute-filepath`
```{r}
manifest_ALL <- parsed %>% 
  pivot_wider(names_from = READ, values_from = V1) %>% 
  mutate(`sample-id` = SAMPLEID,
         `forward-absolute-filepath` = paste(fullpath, r1, sep = ""),
         `reverse-absolute-filepath` = paste(fullpath, r2, sep = "")) 
# View(manifest_ALL)
head(manifest_ALL)
unique(manifest_ALL$SAMPLETYPE)
```

Separate 18S and 16S manifest file. Include controls for both.
```{r}
prokaryote_manifest <- manifest_ALL %>% 
  filter(PRIMER == "16S" | SAMPLETYPE == "control") %>% 
  select(`sample-id`, `forward-absolute-filepath`, `reverse-absolute-filepath`)

eukaryote_manifest <- manifest_ALL %>% 
  filter(PRIMER == "18S" | SAMPLETYPE == "control") %>% 
  select(`sample-id`, `forward-absolute-filepath`, `reverse-absolute-filepath`)
```

```{r}
# Write output as a manifest file
write.table(prokaryote_manifest, file = "manifest-axial-prok", quote=FALSE,col.names=TRUE,row.names=FALSE,sep="\t")

# Write output as a manifest file
write.table(eukaryote_manifest, file = "manifest-axial-euk", quote=FALSE,col.names=TRUE,row.names=FALSE,sep="\t")
```


# QIIME2

Running QIIME2 here: `/scratch/group/hu-lab/qiime-axial-2023_18S_16S`

# Import & compile ASVs

# Set up

```{r}
library(phyloseq)
library(tidyverse)
```

# Import tsv files from QIIME2

```{r}
asv <- read.delim(file = "input-data/samples-asv-table.tsv", sep = "\t", skip = 1)
# head(asv)
```

```{r}
tax <- read.delim(file = "input-data/taxonomy.tsv", sep = "\t")
# head(tax)
```

## Phyloseq import

```{r}
tax_matrix <- tax %>% 
  select(-Consensus) %>% 
  separate(Taxon, into = c("Domain", "Supergroup", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ";", remove = FALSE) %>% 
  column_to_rownames(var = "Feature.ID") %>% 
  as.matrix
```

```{r}
asv_matrix <- asv %>% 
  column_to_rownames(var = "X.OTU.ID") %>% 
  as.matrix
```

Do this with metadata

```{r}
metadata <- read.csv("input-data/metadata_axial-2023_2022.csv")

# head(metadata)
metadata_mod <- metadata %>% 
  column_to_rownames(var = "SAMPLE_NAME")
# metadata_mod

# Import metadata as sample data in phyloseq
samplenames <- sample_data(metadata_mod)
```

Make the phyloseq object

```{r}
# Converting to phylseq objects
tax_phy <- tax_table(tax_matrix)
asv_phy <- otu_table(asv_matrix, taxa_are_rows = TRUE)

# Combine
phy_obj <- phyloseq(asv_phy, tax_phy)

phy_obj <- merge_phyloseq(phy_obj, samplenames)

phy_obj
```

## Make compiled dataframe

```{r}
# head(asv)
# head(tax)
asv_wtax_df <- tax %>% 
  select(-Consensus) %>% 
  separate(Taxon, into = c("Domain", "Supergroup", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ";", remove = FALSE) %>% 
  right_join(asv %>% select(Feature.ID = "X.OTU.ID", everything())) %>% 
  pivot_longer(cols = starts_with("X"), names_to = "FULL_SAMPLE", values_to = "SEQUENCE_COUNT")

head(asv_wtax_df)
unique(asv_wtax_df$FULL_SAMPLE)
```

Isolate plume only samples for preliminary analysis

```{r}
plume_only <- asv_wtax_df %>% 
  filter(SEQUENCE_COUNT > 0) %>% 
  separate(FULL_SAMPLE, into = c("NUMBER", "DIVE", "SITE", "SAMPLE_TYPE", "SAMPLE_NUMBER", "FILTER", "tmp"), sep = "_", remove = FALSE) %>% 
  # Select the samples you want with filer
  filter(grepl("Plume", SAMPLE_NUMBER) | 
           grepl("plume", SAMPLE_NUMBER) |
           SITE == "Background" |
           SITE == "IntlDist" |
           SITE == "ASHES")

head(plume_only)
```

### From plume samples, back to phyloseq

```{r}
tax_matrix_plume <- plume_only %>% 
  select(Feature.ID, Taxon, Domain:Species) %>% 
  distinct() %>% 
  mutate(SUPERGROUP = case_when(
    Phylum == "Alveolata" ~ paste(Phylum, Class),
    Phylum == "Rhizaria" ~ paste(Phylum, Class),
    TRUE ~ Phylum
  )) %>% 
  column_to_rownames(var = "Feature.ID") %>% 
  as.matrix

asv_matrix_plume <- plume_only %>% 
  select(Feature.ID, FULL_SAMPLE, SEQUENCE_COUNT) %>% 
  pivot_wider(names_from = FULL_SAMPLE, values_from = SEQUENCE_COUNT, values_fill = 0) %>% 
  column_to_rownames(var = "Feature.ID") %>% 
  as.matrix
```

```{r}
# metadata <- read.csv("input-data/metadata_axial-2023_2022.csv")
# 
# # head(metadata)
# metadata_mod <- metadata %>% 
#   column_to_rownames(var = "SAMPLE_NAME")
# metadata_mod
```

```{r}

```

Import plume data into phyloseq

```{r}
# Converting to phyloseq objects
tax_phy <- tax_table(tax_matrix_plume)
asv_phy <- otu_table(asv_matrix_plume, taxa_are_rows = TRUE)

# Combine
phy_obj_plume <- phyloseq(asv_phy, tax_phy)
phy_obj_plume

## Combine again with metadata
phy_obj_plume = merge_phyloseq(phy_obj_plume, samplenames)
```

```{r, fig.height=20, fig.width=15}
# plot_heatmap(phy_obj_plume)
```

# QC sequencing run

```{r}
glimpse(asv_wtax_df)
```

```{r, fig.height=20}
table <- asv_wtax_df %>% 
  filter(SEQUENCE_COUNT > 0) %>% 
  group_by(FULL_SAMPLE) %>% 
  summarise(QC_SEQ_SUM = sum(SEQUENCE_COUNT),
            QC_ASV_COUNT = n())
# %>% 
  # pivot_longer(cols = starts_with("QC_"), names_to = "QC", values_to = "VALUE") %>% 
  # ggplot(aes(x = FULL_SAMPLE, y = VALUE)) +
  # geom_bar(stat = "identity", color = "black", aes(fill = QC)) +
  # facet_grid(rows = vars(QC), scales = "free") +
  # # coord_flip() +
  # theme(legend.position = "none",
  #       axis.text.x = element_text(angle = 90))
```

```{r}
# hist(asv_wtax_df$SEQUENCE_COUNT)
```
