---
title: "Deep-sea amplicons"
editor: visual
---

# Set up

```{r}
# | warning: false
library(tidyverse); library(readxl)
library(compositions); library(vegan)
```

# Import tsv files from QIIME2

ASV count table.
```{r}
asv <- read.delim(file = "input-data/deepsea-18s-merged-asv-table-05022025.tsv", sep = "\t", skip = 1)
# head(asv)
colnames(asv)
```

Taxonomy assignment for ASVs.
```{r}
tax <- read.delim(file = "input-data/taxonomy.tsv", sep = "\t")
# head(tax)
```

## Import metadata

Axial Seamount metadata.
```{r}
# Import biotic data
# axial_biotic <- read_excel("metadata-axial-2023.xlsx", sheet = "biotic")
# 
# axial_metadata_samples <- read_excel("metadata-axial-2023.xlsx", sheet = "metadata") %>% 
#   select(-NUM, -OLD_NAME_ARCHIVE) %>% 
#   left_join(axial_biotic) %>% 
#   select(-starts_with("lat_long"), -description_1, -Description_2, -DATE,
#        -primer, -target_population, -lab_num) %>% 
#   add_column(site = "Axial") %>% 
#   mutate(year = case_when(`AXIAL YEAR` == "AXIAL_22" ~ 2022,
#                           `AXIAL YEAR` == "AXIAL_23" ~ 2023),
#          lat = as.numeric(lat),
#          long = as.numeric(long)) %>% 
#   # separate(COORDINATES_mod, c("lat", "long"), sep = ".N.") %>% 
#   select(-`AXIAL YEAR`)

# glimpse(axial_metadata_samples)
# unique(axial_metadata_samples$`AXIAL YEAR`)
```


Mid-Cayman Rise, Gorda Ridge, and other archived samples.

```{r}
# survey_2024 <- read.delim("../microeuks_deepbiosphere_datamine/microeuk-amplicon-survey/data-input/samplelist-metadata.txt", na.strings = "")
# head(survey_2024)
# unique(survey_2024$SAMPLE)
```

```{r}
# head(survey_2024)
# all_metadata <- survey_2024 %>%
#   mutate(COORDINATES_mod = str_replace_all(COORDINATES, " N", ", "),
#          COORDINATES_mod = str_replace_all(COORDINATES_mod, " W", "")) %>%
#   separate(COORDINATES_mod, c("lat", "long"), sep = ", ") %>%
#   mutate(sample_type = case_when(
#            SAMPLETYPE == "Plume" ~ "plume",
#            SAMPLETYPE == "Background" ~ "background",
#            SAMPLETYPE == "Vent" ~ "vent",
#            Sample_or_Control == "Control" ~ "control",
#            SAMPLETYPE == "Microcolonizer" ~ "microcolonizer",
#            SAMPLETYPE == "Incubation" ~ "incubation"),
#          processing = case_when(
#            grepl("Axial_", SAMPLE) ~ "HOG",
#            grepl("GordaRidge", SAMPLE) ~ "SUPR",
#            grepl("_GR_", SAMPLE) ~ "Microcolonizer",
#            grepl("_MCR_", SAMPLE) ~ "HOG"),
#          long = case_when(
#            (SITE == "Axial" | SITE == "GordaRidge") ~ (as.numeric(long) * (-1)),
#            TRUE ~ as.numeric(long)),
#          depth = as.numeric(DEPTH),
#          prok = as.numeric(ProkConc),
#          lat = as.numeric(lat)) %>%
#   add_column(filter = "PES02_sterivex") %>%
#   select(SAMPLE_NAME = SAMPLE, dive_ctd = SAMPLEID,
#          sample_name = VENT, collection_method = Type, sample_type,
#          sample_identifier = Type, filter, processing,
#          location_name = VENT, depth_category = SAMPLETYPE,
#          depth, fluid_origin = Type, lat, long,
#          pH, temp, prok, site = SITE, year = YEAR) %>%
#   add_row(axial_metadata_samples)
# colnames(all_metadata)
```

```{r}
# write.csv(all_metadata, file = "input-data/all_metadata_sep2025.csv", row.names = FALSE)
```


Import and compile all metadata.
Commented out code above is how data was originally combined.

```{r}
all_metadata <- read.csv("input-data/all_metadata_sep2025.csv")
```

```{r}
# colnames(asv)
# colnames(all_metadata)
asv_long_all <- asv %>% 
  pivot_longer(cols = -c("X.OTU.ID"), names_to = "SAMPLE_NAME", values_to = "SEQUENCE_COUNT") %>% 
  select(Feature.ID = "X.OTU.ID", SAMPLE_NAME, SEQUENCE_COUNT) %>% 
  mutate(SAMPLE_NAME = str_replace_all(SAMPLE_NAME, "^X", "")) %>%
  filter(SEQUENCE_COUNT > 0) %>% 
  left_join(all_metadata) #Should join by "SAMPLE_NAME" only.
```

## Curate taxonomy

```{r}
head(tax)
```

Curate taxonomy using the original tax tsv file. 

```{r}
myzo_18s <- c("Perkinsea", "Chrompodellids", "Apicomplexa")
supergroup_list <- c("Stramenopiles", "Rhizaria", "Alveolata", "Opisthokonta", "Ancyromonadida")
supergroup0_list <- c("Archaeplastida", "CRuMs", "Provora", "Cryptista")
##
tax_refine <- tax %>% 
  # select(Taxon) %>% distinct() %>%
  separate(Taxon, into = c("Domain", "Supergroup_0","Supergroup", "Phylum", "Class", "Order" ,"Family", "Genus_spp"), remove = FALSE, sep = ";") %>% 
  filter(Domain == "Eukaryota") %>% 
  mutate(SUPERGROUP_PHYLUM = case_when(
    ## NAs!
    (is.na(Supergroup) & is.na(Supergroup_0)) ~ "Eukaryote-Unannotated",
    (is.na(Supergroup) & !(is.na(Supergroup_0))) ~ paste(Supergroup_0, "Unannotated", sep = "-"),
    (!(is.na(Supergroup)) & is.na(Supergroup_0)) ~ paste(Supergroup, "Unannotated", sep = "-"),
    (is.na(Phylum) & is.na(Supergroup)) ~ paste(Supergroup_0, "Unannotated", sep = "-"),
    Supergroup_0 == "Eukaryota_X" ~ "Eukaryote-Unannotated",
    # Alveolata
    Class == "Syndiniales" ~ paste(Supergroup, Class, sep = "-"),
    Phylum %in% myzo_18s ~ "Alveolata-Myzozoa",
    Phylum == "Dinoflagellata" ~ "Alveolata-Dinophyta",
    (Supergroup == "Alveolata" & grepl("_X", Phylum)) ~ "Alveolata-Unannotated",
    (Supergroup == "Alveolata" & is.na(Phylum)) ~ "Alveolata-Unannotated",
    # Stramenopiles
    (Supergroup == "Stramenopiles" & grepl("_X", Phylum)) ~ "Stramenopiles-Unannotated",
    (Supergroup == "Stramenopiles" & is.na(Phylum)) ~ "Stramenopiles-Unannotated",
    # Rhizaria
    (Supergroup == "Rhizaria" & grepl("_X", Phylum)) ~ "Rhizaria-Unannotated",
    (Supergroup == "Rhizaria" & is.na(Phylum)) ~ "Rhizaria-Unannotated",
    # Amoebozoa
    Supergroup_0 == "Amoebozoa" & is.na(Supergroup) ~ "Amoebozoa-Unannotated",
    Supergroup_0 == "Amoebozoa" ~ paste(Supergroup_0, Supergroup, sep = "-"),
    Supergroup == "Discosea" ~ "Amoebozoa-Lobosa",
    # 
    Supergroup == "Apusomonada" ~ "Apusozoa-Apusomonadida",
    Supergroup == "Breviatea" ~ "Breviatea-Unannotated",
    Supergroup == "Picozoa" ~ "Picozoa",
    Supergroup == "Ancyromonadida" ~ "Ancyromonadida",
    Supergroup_0 == "Excavata" ~ paste("Excavata", Supergroup, sep = "-"),
    Supergroup_0 == "Hacrobia" ~ paste("Hacrobia", Supergroup, sep = "-"),
    #
    Supergroup %in% supergroup_list ~ paste(Supergroup, Phylum, sep = "-"), 
    Supergroup_0 %in% supergroup0_list ~ paste(Supergroup_0, Supergroup, sep = "-"), 
    Supergroup == "Telonemia" ~ "Telonemia"
    ))
  # select(-Genus_spp) %>% distinct()
# unique(tax_refine$SUPERGROUP_PHYLUM)
```
## Create core dfs
```{r}
colnames(asv_long_all)

# All taxa and data
asv_long_tax_all <- asv_long_all %>% 
  left_join(tax) %>% 
  left_join(tax_refine)

colnames(asv_long_tax_all)
```

```{r}
# Taxa, avg across reps, remove NA
asv_long_tax_avg <- asv_long_tax_all %>% 
  group_by(Feature.ID, dive_ctd, sample_name, collection_method, sample_type, sample_identifier, filter, processing, location_name, depth_category, depth, fluid_origin, site, year, lat, long, pH, temp, prok, Taxon, Domain, Supergroup_0, Supergroup, Phylum, Class, Order, Family, Genus_spp, SUPERGROUP_PHYLUM) %>% 
  summarise(MEAN_SEQ = mean(SEQUENCE_COUNT))

unique(asv_long_tax_avg$location_name)
```

# Plot taxonomy

```{r}
unique(asv_long_tax_avg$SUPERGROUP_PHYLUM)

tax_all <- c("#d7b6d7", "#1abc9c", "#34495e", "#a8736f","#125223", "#9b59b6", "#fea02f", "#ff6f61", "#189ab4", "#eb7e17","#f5bfb0", "#de6600", "#007a7a", "#93c47d","#230b36","#c4c168","#a1eb17","#666330","#ee429b",  "#e71414", "#ffe8a4", "#eb655b","#2986cc", "#bcbcbc","#c5cc49","#1b1c1b", "#444444", "#f0f5f1")
```


### Bar plot

```{r, fig.height=13, fig.width=13}
vent_habitat <- c("background", "plume", "vent")
depth_cat <- c("diffuse", "Vent", "peak plume", "Plume", "background", "Background")

df_tax_plotinput <- asv_long_tax_avg %>% 
  #Average across filter types
  group_by(Feature.ID, dive_ctd, sample_name, collection_method, sample_type, location_name, depth_category, depth, site, year, lat, long, pH, temp, prok, SUPERGROUP_PHYLUM) %>% 
  summarise(mean_seq_reps = mean(MEAN_SEQ)) %>% 
  ungroup() %>% 
  filter(sample_type %in% vent_habitat ) %>%
  filter(depth_category %in% depth_cat) %>% 
  filter(location_name != "Soupy background") %>% 
  filter(location_name != "Near vent BW") %>% 
  filter(location_name != "Shallow seawater") %>% 
  filter(!(grepl("Opisthokonta", SUPERGROUP_PHYLUM))) %>%
  filter(!is.na(SUPERGROUP_PHYLUM)) %>%
  group_by(location_name, sample_type, site, year, SUPERGROUP_PHYLUM) %>% 
  summarize(ASV_COUNT = n(),
            SEQUENCE_SUM = sum(mean_seq_reps))


# Plot
df_tax_plotinput %>%
  ggplot(aes(y = location_name, x = SEQUENCE_SUM)) +
  geom_bar(stat = "identity", position = "fill", color = "black", aes(fill = SUPERGROUP_PHYLUM)) +
  scale_fill_manual(values = tax_all) +
  scale_x_continuous(expand = c(0,0)) +
  theme_classic() +
  facet_grid(rows = vars(sample_type, year, site), space = "free", scale = "free") +
  theme(axis.text.x = element_text(color = "black", size = 12),
        axis.text.y = element_text(color = "black", size = 12),
        strip.background = element_blank(),
        strip.text.y = element_text(color = "black", size = 9, angle = 0, hjust = 0, vjust = 0),
        legend.position = "right",
        legend.title = element_blank()) +
  labs(y = "Diffuse vent sites", x = "Relative Sequence Abundance")
# ggsave("figures/vent-nsf-report.svg", width = 9, height = 7, device = "svg", limitsize = FALSE)
```

### ASV distribution

```{r, fig.height=13, fig.width=13}
head(df_tax_plotinput)
df_tax_plotinput %>% 
  ggplot(aes(y = location_name, x = SUPERGROUP_PHYLUM, size = ASV_COUNT, fill = SUPERGROUP_PHYLUM)) +
    # geom_tile(stat = "identity", color = "black") +
  geom_point(shape = 21, color = "black") +
  scale_fill_manual(values = tax_all) +
  # scale_size_manual(values = c(1,8)) +
  scale_x_discrete(position = "top") +
  theme_classic() +
  facet_grid(rows = vars(sample_type, year, site), space = "free", scale = "free") +
  theme(axis.text.x = element_text(color = "black", size = 12, angle =90, hjust = 0, vjust = 0),
        axis.text.y = element_text(color = "black", size = 12),
        strip.background = element_blank(),
        strip.text.y = element_text(color = "black", size = 9, angle = 0, hjust = 0, vjust = 0),
        legend.position = "none",
        legend.title = element_blank()) +
  labs(y = "Diffuse vent sites", x = "Relative Sequence Abundance")
# ggsave("figures/vent-nsf-report.svg", width = 9, height = 7, device = "svg", limitsize = FALSE)
```


## Heatmap

```{r}
unique(asv_long_tax_avg$year)
yrs <- c(2019, 2020, 2023)

df_tax_matrix <- asv_long_tax_avg %>% 
  #Average across filter types
  group_by(Feature.ID, dive_ctd, sample_name, collection_method, sample_type, location_name, depth_category, depth, site, year, lat, long, pH, temp, prok, SUPERGROUP_PHYLUM) %>% 
  summarise(mean_seq_reps = mean(MEAN_SEQ)) %>% 
  ungroup() %>% 
  filter(sample_type %in% vent_habitat ) %>%
  filter(depth_category %in% depth_cat) %>% 
  filter(location_name != "Soupy background") %>% 
  filter(location_name != "Near vent BW") %>% 
  filter(location_name != "Shallow seawater") %>% 
  filter(!(grepl("Opisthokonta", SUPERGROUP_PHYLUM))) %>%
  filter(!is.na(SUPERGROUP_PHYLUM)) %>%
  filter(year %in% yrs) %>% 
  unite(SAMPLE_INFO, location_name, sample_type, site, sep = "_") %>% 
  group_by(SAMPLE_INFO, SUPERGROUP_PHYLUM) %>% 
  summarize(SEQUENCE_SUM = sum(mean_seq_reps)) %>% 
  pivot_wider(names_from = SUPERGROUP_PHYLUM, values_from = SEQUENCE_SUM, values_fill = 0) %>% 
  column_to_rownames(var = "SAMPLE_INFO")
```
Data transformation.
```{r}
# ?decostand
df_tax_norm <- vegan::decostand(df_tax_matrix, method = "standardize", MARGIN = 1)
# head(df_tax_norm)
```


```{r, fig.height=13, fig.width=13}
df_tax_norm %>% 
  rownames_to_column(var = "SAMPLE_NAME") %>% 
  pivot_longer(cols = -c(SAMPLE_NAME), names_to = "SUPERGROUP_PHYLUM", values_to = "norm_seq") %>% 
  separate(SAMPLE_NAME, into = c("location_name", "sample_type", "site"), sep = "_", remove = FALSE) %>% 
  ggplot(aes(y = location_name, x = SUPERGROUP_PHYLUM, fill = norm_seq)) +
    geom_tile(stat = "identity", color = "black") +
  scale_fill_fermenter(type = "seq", palette = "Reds", direction = 1) +
  scale_x_discrete(position = "top") +
  theme_classic() +
  facet_grid(rows = vars(sample_type, site), space = "free", scale = "free") +
  theme(axis.text.x = element_text(color = "black", size = 12, angle =90, hjust = 0, vjust = 0),
        axis.text.y = element_text(color = "black", size = 12),
        strip.background = element_blank(),
        strip.text.y = element_text(color = "black", size = 9, angle = 0, hjust = 0, vjust = 0),
        legend.position = "right",
        legend.title = element_blank()) +
  labs(y = "Diffuse vent sites", x = "Relative Sequence Abundance")
# ggsave("figures/vent-nsf-report.svg", width = 9, height = 7, device = "svg", limitsize = FALSE)
```


# Diversity

```{r}
yrs <- c(2019, 2020, 2023)

df_all_data_matrix <- asv_long_tax_avg %>% 
  #Average across filter types
  group_by(Feature.ID, dive_ctd, sample_name, collection_method, sample_type, location_name, depth_category, depth, site, year, lat, long, pH, temp, prok, SUPERGROUP_PHYLUM) %>% 
  summarise(mean_seq_reps = mean(MEAN_SEQ)) %>% 
  ungroup() %>% 
  filter(sample_type %in% vent_habitat ) %>%
  filter(depth_category %in% depth_cat) %>% 
  filter(location_name != "Soupy background") %>% 
  filter(location_name != "Near vent BW") %>% 
  filter(location_name != "Shallow seawater") %>% 
  filter(!(grepl("Opisthokonta", SUPERGROUP_PHYLUM))) %>%
  filter(!is.na(SUPERGROUP_PHYLUM)) %>%
  filter(year %in% yrs) %>% 
  unite(SAMPLE_INFO, location_name, sample_type, site, sep = "_") %>% 
  group_by(SAMPLE_INFO, Feature.ID) %>% 
  summarize(SEQUENCE_SUM = sum(mean_seq_reps)) %>% 
  pivot_wider(names_from = SAMPLE_INFO, values_from = SEQUENCE_SUM, values_fill = 0) %>% 
  column_to_rownames(var = "Feature.ID")
```

```{r}
shannon <- vegan::diversity(df_all_data_matrix, index = "shannon", MARGIN = 2)
# head(shannon)
```
```{r}
as.data.frame(shannon) %>% 
  rownames_to_column(var = "SAMPLE_NAME") %>% 
  separate(SAMPLE_NAME, into = c("location_name", "sample_type", "site"), sep = "_", remove = FALSE) %>% 
  mutate(SAMPLE_TYPE = case_when(
    sample_type == "vent" ~ "Vent",
    sample_type == "plume" ~ "Plume",
    sample_type == "background" ~ "Background"
  )) %>% 
  ggplot(aes(y = SAMPLE_TYPE, x = shannon)) +
  geom_boxplot(color = "black", width = 0.3) +
    geom_jitter(stat = "identity", color = "black", width = 0.1, size = 3,
                aes(fill = site, shape = site)) +
  scale_shape_manual(values = c(21, 22, 23, 24)) +
  scale_fill_manual(values = c("#28a87e", "#111f61", "#bf2c7d", "#d6be47")) +
  theme_classic() +
  theme(axis.text.x = element_text(color = "black", size = 12),
        axis.text.y = element_text(color = "black", size = 12),
        strip.background = element_blank(),
        strip.text.y = element_text(color = "black", size = 9, angle = 0, hjust = 0, vjust = 0),
        legend.position = "right",
        legend.title = element_blank(),
        panel.grid.major = element_line()) +
  labs(y = "", x = "Shannon Diversity")
```

```{r}
# ?diversity
invsimp <- vegan::diversity(df_all_data_matrix, index = "invsimpson", MARGIN = 2)
# head(invsimp)
```

```{r}
as.data.frame(invsimp) %>% 
  rownames_to_column(var = "SAMPLE_NAME") %>% 
  separate(SAMPLE_NAME, into = c("location_name", "sample_type", "site"), sep = "_", remove = FALSE) %>% 
  mutate(SAMPLE_TYPE = case_when(
    sample_type == "vent" ~ "Vent",
    sample_type == "plume" ~ "Plume",
    sample_type == "background" ~ "Background"
  )) %>% 
  ggplot(aes(y = SAMPLE_TYPE, x = shannon)) +
  geom_boxplot(color = "black", width = 0.3) +
    geom_jitter(stat = "identity", color = "black", width = 0.1, size = 3,
                aes(fill = site, shape = site)) +
  scale_shape_manual(values = c(21, 22, 23, 24)) +
  scale_fill_manual(values = c("#28a87e", "#111f61", "#bf2c7d", "#d6be47")) +
  theme_classic() +
  theme(axis.text.x = element_text(color = "black", size = 12),
        axis.text.y = element_text(color = "black", size = 12),
        strip.background = element_blank(),
        strip.text.y = element_text(color = "black", size = 9, angle = 0, hjust = 0, vjust = 0),
        legend.position = "right",
        legend.title = element_blank(),
        panel.grid.major = element_line()) +
  labs(y = "", x = "Inverse Simpson")
```
